\plaintitle

### 2017-04-27 statistical-quorum.tex

\begin{table*}
  \centering
  \begin{tabular}{l r r r r r r r}
    {\small\textit{Vote}} & \multicolumn{2}{r}{\small \textit{First option}} & \multicolumn{2}{r}{\small \textit{Second option}} & {\small \textit{Total eligible voters}} & {\small \textit{Turnout}} & {\small \textit{Decision credibility}} \\
    \midrule
    Brexit & 17,410,742 & $51.9\%$ & 16,141,241 & $48.1\%$ & 46,500,001 & $72.2\%$ & $59.8\%$ \\
    Presidential elections & 62,984,825 & $48.9\%$ & 65,853,516 & $51.1\%$ & 230,585,915 & $55.9\%$ & $52.6\%$ \\
  \end{tabular}
  \caption{Votes cast in Brexit, a 2016 referendum for United Kingdom's withdrawal from the European Union~\protect\cite{brexitresults},
  and 2016 USA presidential elections, popular vote~\protect\cite{usaresults1,usaresults2}. Simplified to only two options and without
  invalid votes. The last column shows computed decision credibility as described in Section~\ref{sec:statistical-quorum}. We see that
  credibility is low, which we can mostly attribute to the fact that there is small difference in votes between both options so
  with more votes any option could end up winning.}\label{tab:example}
\end{table*}

To answer this question we are proposing a novel metric, \emph{decision credibility}, which uses statistical
hypothesis testing to evaluate the credibility of the result.
It uses distribution of votes made to compute the probability of decision passing if everyone voted,
and not just those who voted.
Intuitively, if in a community of 100 members 40 members vote for a decision to pass and nobody against,
we can see that if everyone voted, the probability of majority of population voting for a decision is high.
On the other hand, if 41 members for in favor and 40 against, we cannot really say much about what would a decision
be if everyone voted.
Only two more votes against can completely change the outcome.
We describe decision credibility in more detail in Section~\ref{sec:statistical-quorum}\todo{Hm, SIGCGI template seems to not display section numbers. So how useful is to reference a section number?}.

Based on decision credibility we can define a \emph{statistical quorum} as a threshold which a decision's credibility
has to pass for a decision to be seen as legitimate.
For example, if we adopt $90\%$ as a reasonable requirement for statistical quorum and we compute decision
credibility for votes in Table~\ref{tab:example} we see that in both cases we do not reach statistical quorum
which puts legitimacy of those decisions into a question.
We discuss a way for a community to determine a reasonable statistical quorum in Section~\ref{sec:discussion}.

Traditionally it is argued that legitimacy comes from the fact that members had a chance to participate
and they decided not to, giving legitimacy to make a decision to those who did participate.
We argue that this is a problematic view because there are many restrictions on why members could not
participate and vote.
For example, they might not had time to participate because of other commitments, or motivation, or they might determined
that they are not informed enough to participate.
Moreover, members are from different reasons in their communities and potentially contribute in other ways
than by participating in its democratic processes.
Instead of assuming that everyone could participate if they would want, or trying to increase participation,
we can see amount of participation members are able to do as a limited resource, a \emph{collective attention}
a community has available to utilize.

We can compute decision credibility during ongoing voting process to give feedback to the community
on how contentious a proposed decision is how much more votes are necessary for a statistical quorum
to be reached.
This allows the community to focus its collective attention towards decisions which require more votes
to determine a legitimate result.
Instead of trying to increase participation on all democratic decisions being made, a community can
do that selectively through an automatic feedback loop.

To evaluate the use of statistical quorum in decision making of democratic communities we have developed
a web application PeerMind which integrates online discussion with online voting.
The application exposes computed decision credibility to users in real-time and through that provides
the feedback loop needed for improving utilization of collective attention.
We present the app in Section~\ref{sec:peermind}.

@@

\todo[inline]{Structure of the rest of the paper.}

@@

% threshold and quorum at https://www.publications.parliament.uk/pa/ld200910/ldselect/ldconst/99/99.pdf
% how does some claim that quorum are in fact bad, and counter democratic, but do not offer a solution how to evaluate them

@@

@@

% describe why 0.90% is a reasonable statistical quorum and how to determine one for your community

@@

\bibliographystyle{sigchi}
\bibliography{refs}

>>>>

\begin{table*}
  \centering
  \begin{tabular}{l r r r r r r r}
    {\small\textit{Vote}} & \multicolumn{2}{r}{\small \textit{First option}} & \multicolumn{2}{r}{\small \textit{Second option}} & {\small \textit{Total eligible voters}} & {\small \textit{Turnout}} & {\small \textit{Decision credibility}} \\
    \midrule
    Brexit & 17,410,742 & $51.9\%$ & 16,141,241 & $48.1\%$ & 46,500,001 & $72.2\%$ & $59.8\%$ \\
    Presidential elections & 62,984,825 & $48.9\%$ & 65,853,516 & $51.1\%$ & 230,585,915 & $55.9\%$ & $52.6\%$ \\
  \end{tabular}
  \caption{Votes cast in Brexit, a 2016 referendum for United Kingdom's withdrawal from the European Union~\protect\cite{brexitresults},
  and 2016 USA presidential elections, popular vote~\protect\cite{usaresults1,usaresults2}. Simplified to only two options and without
  invalid votes. The last column shows computed decision credibility as described in Section \secref{sec:statistical-quorum}. We see that
  credibility is low, which we can mostly attribute to the fact that there is small difference in votes between both options so
  with more votes any option could end up winning.}\label{tab:example}
\end{table*}

To answer this question we are proposing a novel metric, \emph{decision credibility}, which uses statistical
hypothesis testing to evaluate the credibility of the voting result.
It uses distribution of votes made to evaluate the probability of decision passing if everyone actually voted,
and not just those who voted.
Intuitively, if in a community of 100 members 40 members vote for a decision to pass and nobody against,
we can see that if everyone voted, the probability of majority of population voting for a decision is high.
On the other hand, if 41 members for in favor and 40 against, we cannot really say much about what would a decision
be if everyone voted.
Only two more votes against can completely change the outcome.
Hence, the suggested action would be for the community to gather more votes and continue discussion around the decision.
We describe decision credibility in more detail in Section \secref{sec:statistical-quorum}.

Based on decision credibility we can define a \emph{statistical quorum} as a threshold which a decision's credibility
has to pass for a decision to be seen as legitimate.
For example, if we adopt $90\%$ as a reasonable requirement for statistical quorum and we compute decision
credibility for votes in Table~\ref{tab:example} we see that in both cases we do not reach statistical quorum
which puts legitimacy of those decisions into a question.
We discuss a way for a community to determine a reasonable statistical quorum in Section \secref{sec:discussion}.

Traditionally it is argued that legitimacy comes from the fact that members had a chance to participate
and they decided not to, giving legitimacy to make a decision to those who did participate.
We argue that this is a problematic view because there are many restrictions on why members could not
participate and vote.
For example, they might not had time to participate because of other commitments, or motivation, or they might determined
that they are not informed enough to participate.
Moreover, members are from different reasons in their communities and potentially contribute in other ways
than by participating in its democratic processes.
Instead of assuming that everyone could participate if they would want, or trying to increase participation,
we can see amount of participation members are able to do as a limited resource, a \emph{collective attention}
a community has available to utilize for its decision making.

We can compute decision credibility during ongoing voting process to give feedback to the community
on how contentious a proposed decision is how much more votes are necessary for a statistical quorum
to be reached.
This allows the community to focus its collective attention towards decisions which require more votes
to determine a legitimate result, while reducing attention towards proposals which already have overwhelming support.
Instead of trying to increase participation on all democratic decisions being made, a community can
do that selectively through an automatic feedback loop, trying to reach statistical quorum for their decisions.

We built an open source web application PeerMind integrating statistical quorum with online voting.
The application exposes computed decision credibility to users in real-time and through that provides
the feedback loop needed for improving utilization of collective attention.
We present the application in Section \secref{sec:peermind} and use it for evaluating statistical quorum in
Section \secref{sec:evaluation}.

@@

\todo[inline]{Repeat the structure of the rest of the paper? Is there a better way to reference sections than
referencing it by name?}

@@

\todo[inline]{Repeat references from "There have been many improvements proposed to this basic form and various analysis made."?}

\todo[inline]{Threshold and quorum at \url{https://www.publications.parliament.uk/pa/ld200910/ldselect/ldconst/99/99.pdf}.
How do some claim that quorum are in fact bad, and counter democratic, but do not offer a solution how to evaluate decisions.}

\todo[inline]{Voting auditing work.}

\todo[inline]{Delegation as way to increase legitimacy, Liquid Democracy, etc.}

@@

We evaluate statistical quorum in two ways. First, we evaluate it theoretically and explore its mathematical properties.
Then we evaluate it through real-world use of the PeerMind application by a 140 members community.

\subsection{Properties of Statistical Quorum}

\todo[inline]{Theoretical evaluation, diagrams. Get readers to understand some of its properties visually.}

\subsection{Use in Existing Democratic Community}

% imperial: running it for x weeks in student community which has traditionally weekly meeting
% issues with quorum
% participation of everyone
% they used online app for x weeks
% we compare their old minutes with performance of new system

% interviews and surveys

\todo[inline]{Make sure that we are supporting both of our claims well: this statistical quorum is a good metric and
collective attention can be utilized better.}

PeerMind source code is available under open source licence at \url{http://redacted.for.blind.review} and
anonymized dataset of interactions with the app, including votes, is available at \url{http://redacted.for.blind.review}.

@@

\todo[inline]{Describe why is $0.90\%$ a reasonable statistical quorum and how to determine one for your community.}

\todo[inline]{Describe interesting example how we had a bug which were showing a very low statistical quorum for a failing
vote and so the whole community very engaged for a whole month to decide and resolve it. So how they trusted the
app and the metric and how it was really able to direct the attention of the community. It seems it works even if it is
buggy or not perfect, because it has worked so well in the past that members got used to the fact that when it was saying
that there is a consensus or not on an issue that that was also their own read of the community. So for the purpose of
directing attention the metric does not have to mathematically perfect.}

\todo[inline]{Security. How to prevent a group of people to vote and by trick reaching statistical quorum.
We use dynamic statistical quorum to address this, but one other approach is also for voting to stay open and
decisions can then be in the future revoked by people adding more votes, or changing them. But for some decisions
this might not really be possible, because decisions might not be reversible. You can revoke a law, but if you built
a building as a consequence of a decision?}

\section{Future Work}

\todo[inline]{More work on security evaluation of potential attacks when a community is using statistical quorum to force a decision.
Is there a game theoretical approach to it?}

\todo[inline]{How to surface information about credibility to users better than just a number? Visualizations? How can we help them
understand what exactly this is and how the math works?}

@@

% REFERENCES FORMAT
% References must be the same font size as other body text.
\bibliographystyle{sigchi}
\bibliography{refs}
